Towards Semantic Patches

### 2010-06-18 papers/MIPS2010/abstract.tex
* Preprocesseur.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@100 403e655e-980a-47bd-8c60-1ee5348bb6b7
* Quelques commentaires.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@101 403e655e-980a-47bd-8c60-1ee5348bb6b7

\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
\usepackage[pdftex,backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{fancyhdr}
\input xy
\xyoption{all}
\usepackage[protrusion=true,expansion=true]{microtype}

@@

\title{Semantic Patches
  % pas sûr que le nom soit encore adapté
}
\author{Matthias Puech $\quad\Leftrightarrow\quad$ Yann Régis-Gianas}
\date{}

\begin{document}
\maketitle

@@

It is interesting to investigate the possible relationship between the
very \emph{methodologies} or even \emph{daily workflow} employed in
both fields of mathematics and computer science: one trying to prove a
theorem, the other constructing a program to fulfill a task. How does
one or the other elaborates his object of study? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or the other
way around, how do one rely on existing work to build up new results?
Finally, how to relate how both scientists collaborate in a team?

Recent efforts in the formalization of mathematical results have
naturally led to these questions and many of them remain largely
unanswered, but the tendency seems to be to adapt existing methods
coming from software development, witness for example the recent
introduction of modules, file-based scripts and separate compilation
in proof assistants like \textsf{Coq} or \textsf{Matita}, the use of
dependency management tools (\textsf{make}) or version control system
(\textsf{git}, \textsf{svn}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

As intelligent and widely adopted as this tool chain has gotten, we
believe that they are not adapted for the new, demanding requirements
of proof developments. Indeed, whereas compilation of a program is
usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery process by
\cite{lakatos1964proofs}, and we believe that they should be taken
into account in the realization of mathematical assistants like
\textsf{Coq} or \textsf{Matita}.

@@

In the light of the propositions-as-types paradigms where
proof-checking boils down to type-checking, we abstract from whether
we are talking about proofs or programs, so that our development could
eventually also be used in traditional software development.

\section{Methodologies}

We propose here to devise a system for \emph{semantic patches}. It
% TODO changer le nom?
substitutes the idea of textual transformation by:
\begin{itemize}
\item First preferring an abstract syntax tree representation instead
  of plain text;
\item Secondly embedding semantic, or typing data into the
  transformations, in order to be able to reason on them.
\item Finally, perform type-checking in an incremental way, that is
  type only the syntactic difference and reuse derivation for the
  already known subprograms.
\end{itemize}

We want to design a \emph{language of patches}, able to represent
% TODO changer le nom?
these transformations and their semantic properties, thus capturing
local changes in programs, as well as their global effect on the whole
project. We can see this goal as a refinement of the former idea of
``dependency''.

\subsection{Data \emph{vs.} patch-oriented}

\subsection{Related Works}

\section{A kernel for incremental type-checking}

@@

@@

@@

>>>>

\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
\usepackage[backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{fancyhdr}
\usepackage[all,cmtip]{xy}
%\usepackage[protrusion=true,expansion=true]{microtype}

@@

% Commandes à désactiver avant la soumission.
% Remarques sur la structure.
\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
% Remarques sur le contenu.
\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}

\title{Towards Semantic Patches}
  % pas sûr que le nom soit encore adapté

\author{Matthias Puech $\quad\Leftrightarrow\quad$ Yann Régis-Gianas}
\date{}

\begin{document}

\maketitle

@@

\remplan{Motivations générales}

It may be fruitful to investigate the possible relationship between the
very \emph{methodologies} or even \emph{daily workflow} employed in
both fields of mathematics and computer science: one trying to prove a
theorem, the other constructing a program to fulfill a task. How does
one or the other elaborates his object of study? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or the other
way around, how do one rely on existing work to build up new results?
% Y: Je ne comprends pas tout à fait la dernière question: 
Finally, how to relate how both scientists collaborate in a team?

Recent efforts in the formalization of mathematical results have
naturally led to these questions and many of them remain largely
unanswered, but the tendency seems to be to adapt existing methods
coming from software development, as illustrated for example by the
recent introduction of modules, file-based scripts and separate
compilation in proof assistants like \textsf{Coq}~{\cite{CoqDocWeb}}
or \textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
dependency management tools (\textsf{Make}) or version control system
(\textsf{GIT}, \textsf{Subversion}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

As intelligent and widely adopted as this tool chain has gotten, we
believe that they are not adapted for the new, demanding requirements
of proof developments. Indeed, whereas compilation of a program is
usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery
process~\cite{lakatos1964proofs}, and we believe that they should be
taken into account in the realization of mathematical assistants.
% \textsf{Coq} or \textsf{Matita}.

\remtext{Fri Jun 18, 2010 9:55. AM Ici, on pourrait rajouter un
  paragraphe pour dire qu'il existe des travaux sur la formalisation
  de metatheories dans les assistants de preuve, que ce soit en TWELF
  ou Coq. Puis, on explicite notre but général qui est de construire
  un système/framework qui remplace la chaîne d'outils traditionnelle
  en important/adaptant la technologie issue de la formalisation de
  metatheories.}

\remplan{Sujet du papier : on focalise sur un problème bien précis}

\remtext{Fri Jun 18, 2010 10:00 AM.
  Un peu de liant. Pour arriver à notre but, il y a une
  question importante à résoudre: si on te donne une méta-théorie
  formalisée, es-tu capable de décrire l'historique d'un développement
  dans cette théorie? Cela induit deux questions : dans quel langage
  et est-ce que ce langage est utilisable (parce qu'a priori, ce n'est
  pas gagné car le développeur travaille en mode ``instantané'' et ne
  veut pas se trimballer ses casseroles tout le temps\ldots).}

@@

\remtext{Fri Jun 18, 2010 10:06 AM.  Le paragraphe précédent n'est pas
  assez précis je pense et en même temps en dit trop. À la place, je
  propose d'énoncer succintement les contributions de ce papier puis 
  le plan. Ce serait quelque chose comme:
  1. A description language for semantic repositories.
  2. Theory development through semantic patches. 
  3. Related work
  4. Further work
}

\remplan{Fri Jun 18, 2010 10:08 AM. Du coup, je restructure un peu la suite.}

\section{A kernel language to describe semantic repositories}

\subsection{How to describe a development?}

\remplan{Fri Jun 18, 2010 10:09 AM. Pour commencer, on peut essayer 
  de répondre à la question: ``À quoi sert ce langage?''. }

\remtext{Fri Jun 18, 2010 10:10 AM. Il sert à décrire des objets d'une
  théorie et, surtout, des transformations de ces objets (valides) en
  de nouveaux objets (valides). Il sert donc à décrire des applications
  de méta-théorèmes.}

\remplan{Fri Jun 18, 2010 10:13 AM. Ensuite, est-ce qu'il n'existe pas déjà
  des langages adaptés à ça? }

\remtext{Fri Jun 18, 2010 10:15 AM. Pas tout à fait. Certes, LF
  fournit un cadre qui autorise la représentation des metathéories et
  des objets, construits par application de règles
  admissibles/métathéorèmes. Cependant, la syntaxe de ces objets 
  est statique/extensionnel~: on décrit l'objet comme le résultat
  de sa construction en oubliant son histoire. Pour représenter 
  l'historique d'un développement, notre langage
  doit au contraire maintenir l'ordre dans lequel les objets ont
  été construits ainsi que la façon dont ils l'ont été.}

\remtext{Fri Jun 18, 2010 10:26 AM. Ce n'est pas peut être pas tout à fait vrai que personne n'a
  fait de proposition de langages qui décrivent un développement ...
  On peut peut-être citer Pfenning, Klint and co et faire une référence
  en avant vers le related work.}

\remtext{Fri Jun 18, 2010 10:20 AM. Remarque annexe: je pense qu'à ce
  stade de l'explication, il n'est pas nécessaire de parler
  d'incrémentalité et de partage maximal. Je crois que c'est une
  conséquence pratique heureuse de notre formalisation mais qui ne
  fait pas réellement partie des nécessités/besoins initiaux.
  Ce que je veux dire, c'est que l'incrémentalité est une propriété
  orthogonale à la capacité de dénoter une historique.}

@@

\remtext{Fri Jun 18, 2010 10:26 AM. 
  Du coup, il faudrait reformuler un peu les paragraphes précédents.}

\subsection{Formalization}

\remplan{On rentre dans les détails techniques. Je crois qu'il faut
  prendre un peu de temps pour définir précisément les termes utilisés
  (repository, patch, etc) et expliquer/paraphraser les règles.}

@@

% Y: Remarque en passant: pour maintenir des notations cohérentes, il
% vaut mieux se définir des macros pour absolument tout
% (metavariables, jugements, ...). Ca passe mieux à l'échelle.  On
% verra ca pour le prochain papier :-).

@@

\section{Development through semantic patches}

\remplan{Fri Jun 18, 2010 10:50 AM. L'objectif ici, c'est de décrire
  une (possible) architecture pour rendre utilisable le langage 
  décrit dans la section précédente. En gros, on énumère les 
  opérations qui interviennent dans l'utilisation et on les décrit
  informellement à l'aide d'un exemple.}

\remtext{Fri Jun 18, 2010 10:52 AM. Par contre, je ne sais pas quel
  exemple utilisé. Le STLC? Est-ce que ça fait ``assez'' système
  logique pour ce workshop?}

\subsection{A layered architecture}

% À mettre dans une figure. 
\begin{figure}
\begin{center}
% FIXME: il y a un problème avec la flèche double phi/psi.
\xymatrix{
\textrm{History of metatheory objects}
&R 
\ar@{->}[d]_{\phi}
& \Delta \ar@{->}[r]^{\theta(R)} & R' \\
\textrm{Metatheory objects} 
&t \ar@{->}[d]^{\pi}
\ar@{->}[u]_{\psi}
& t' \ar@{->}[u]_{\nabla(t)}\\
\textrm{Concrete views on metatheory objects} 
&w \ar@{.>}[r] & w' \ar@{->}[u]_{\sigma(t, w)}
}
\end{center}
\label{fig:architecture}
\caption{A layered architecture for a semantic patch framework.}
\end{figure}

\remtext{Fri Jun 18, 2010  1:32 PM. Ici, on décrit la figure en suivant
  le modèle de flot du système. On décrit le rôle des fonctions, 
les problèmes posés.}

\paragraph{From an abstract proof tree to a concrete view, and back}
\-

\remtext{Fri Jun 18, 2010  1:36 PM. Un arbre de preuves, ce n'est
  pas très pratique. Pourtant, c'est moralement sur cet objet 
  que s'opère le développement et non sur le code source, qui
  n'en est qu'une représentation. On justifie la nécessité de
  projection et on fait référence aux lentilles de Pierce pour
  expliquer $\sigma$ et $\pi$.}

\remtext{Fri Jun 18, 2010 1:46 PM. Autre point important sur les vues
  concrètes~: elles peuvent servir à parler de l'historique.  Voir le
  module MetaAST de l'implémentation. Ces vues sont utiles par exemple
  lorsque l'utilisateur doit construire un patch d'adaptation. Il a
  alors toute la puissance des métathéorèmes ainsi que tous les objets
  déjà construits pour décrire son patch.}

\paragraph{Capturing the evolution of abstract proof trees}
\-

\remtext{Fri Jun 18, 2010 1:38 PM. Il faut différencier deux arbres de
  syntaxe. Une méthode : la distance d'édition syntaxique permet de
  déduire des applications des règles d'élimination et d'introduction
  que l'on peut utiliser pour construire $\Delta$. Référence sur 
  la littérature sur les distances d'édition entre arbres. Discussion
  sur l'élaboration de patchs de haut niveau (comme le renommage etc). }

\paragraph{Propagating changes through dependencies}
\-

\remtext{Fri Jun 18, 2010 1:42 PM. Si on se contente de concaténer
  $\Delta$ à $R$, on n'a intégré qu'une nouvelle version de l'objet
  initial~$t$ dans $R$ mais on n'a pas propagé ces modifications
  auprès des clients de $t$. Cette propagation n'est \textit{a priori}
  pas triviale~: rien ne dit que $t'$ à un sens équivalent à $t$. Il
  faut donc adapter les clients en demandant un patch d'adaptation à
  l'utilisateur. On présente ces demandes sous la forme de patch à
  compléter (on fournit un peu de contexte et on a un trou pour y
  injecter une nouvelle sous-dérivation mais on peut aussi modifier le
  contexte pour l'adapter). La construction d'un patch d'adaptation
  est un processus interactif~: on ne sait pas a priori borner le
  contexte nécessaire à la construction du patchs d'adaptation donc on
  peut imaginer un sorte de jeu où le système fait augmenter le
  contexte (en suivant les dépendances) si l'utilisateur ne sait pas
  construire le patch d'adaptation. En passant, quand on parle d'utilisateur
  ici, ce peut être aussi bien un programme, comme un type-checker
  pour le langage objet, ou bien un programmeur.}

\subsection{Application on simply typed $\lambda$-calculus}

\section{Related work}

\remtext{Dolphin, Twelf, etc}

\remtext{Semantic patches à la coccinnelle}

\section{Further work}

\remtext{Extension vers une version constructiviste?}

\remtext{Algorithme d'élaboration de patchs?}

\remtext{Parler du bootstraping?}

\section{Conclusion}

%% In the light of the propositions-as-types paradigms where
%% proof-checking boils down to type-checking, we abstract from whether
%% we are talking about proofs or programs, so that our development could
%% eventually also be used in traditional software development.

%% \section{Methodologies}

%% We propose here to devise a system for \emph{semantic patches}. It
%% % TODO changer le nom?
%% substitutes the idea of textual transformation by:
%% \begin{itemize}
%% \item First preferring an abstract syntax tree representation instead
%%   of plain text;
%% \item Secondly embedding semantic, or typing data into the
%%   transformations, in order to be able to reason on them.
%% \item Finally, perform type-checking in an incremental way, that is
%%   type only the syntactic difference and reuse derivation for the
%%   already known subprograms.
%% \end{itemize}

%% We want to design a \emph{language of patches}, able to represent
%% % TODO changer le nom?
%% these transformations and their semantic properties, thus capturing
%% local syntactic changes in programs (the abstract syntax tree), as
%% well as their global effect on the whole project (the typing
%% derivations). We can see this goal as a refinement of the former idea
%% of ``dependency''.
